---
title: "Supplemental Material"
subtitle: ""
author: Mia Carmichael
institute: "University of Georgia"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: article
output:
  bookdown::word_document2: 
    toc: false
bibliography: ../references.bib
csl: ../apa.csl
---

```{r, echo=FALSE, warning=FALSE}
library(dplyr)
library(pander)
library(knitr)
library(here)
library(car)
library(ggplot2)
library(broom)

phyladata<- readRDS(here("././data/processed_data/processeddata_phyladata.rds"))
phylatest <- readRDS(here("././data/processed_data/processeddata_phyladata.rds"))
phylacomb <- readRDS(here("././results/phylacomb.rds"))

phyladata2<- readRDS(here("././data/processed_data/processeddata_phyladata2.rds"))
phylatest2 <- readRDS(here("././data/processed_data/processeddata_phyladata2.rds"))
phylacomb2 <- readRDS(here("././results/phylacomb.rds"))

```


## Bivariate Analysis
Bacterial abundance verses time period. ANOVA allows us to see that there is variance between the bacterial abundance at beginning and end of the feedlot period in High-RFI steers (Table 5, Table 6). This indicates that the microbiome did experieience changes over the course of the feedlot period.

Looking at the correlation, we can see that there is a high positive correlation between the change from the beginning of the feedlot period to the end of the feeddlot period for phyla,  and therefore the correlation between them will be closer to 1.

# Looking at ANOVA
```{r, warning=FALSE, echo=FALSE}
# High
plot(Abundance ~ Phyla, data = phylacomb)
phylahigh.aov <- aov(Abundance ~ Phyla, data = phylacomb)
summary(phylahigh.aov)
phylahigh.aovsummary <- summary(phylahigh.aov)
print(phylahigh.aovsummary)

saveRDS(phylahigh.aovsummary, file = here("./results/phylahigh_aov_table.rds"))



# Low
plot(Abundance ~ Phyla, data = phylacomb2)
phylalow.aov2 <- aov(Abundance ~ Phyla, data = phylacomb2)
summary(phylalow.aov2)
phylalow.aovsummary2 <- summary(phylalow.aov2)
print(phylalow.aovsummary2)

saveRDS(phylalow.aovsummary2, file = here("./results/phyla_low_aov_table.rds"))




```


# Looking at a correlation between data points at the beginning and end of feedlot period
```{r, warning=FALSE, echo=FALSE}

#High
phylacor <- cor(phylatest$Beginning, phylatest$End)

phylacor.table <-broom::tidy(phylacor)
phylacor.table
saveRDS(phylacor.table, file = here("./results/phylahightablecor.rds"))


#Low
phylacor2 <- cor(phylatest2$Beginning, phylatest2$End)

phylacor2.table <-broom::tidy(phylacor2)
phylacor2.table
saveRDS(phylacor2.table, file = here("./results/phylahightablecor2.rds"))


```


# Looking at Linear Model
```{r, echo=FALSE, warning=FALSE}

# High
linearMod.1 <- lm(Beginning ~ End, data = phylatest)
print(linearMod.1)
summary(linearMod.1)

Anova(linearMod.1, type = "III")
modelSummary.1 <- summary(linearMod.1)
print(modelSummary.1)

# Place results back into a data frame
linearMod.1table <-broom::tidy(modelSummary.1)
linearMod.1table

# Save chart with p-value
saveRDS(linearMod.1table, file = here ("././results/linearMod_1table.rds"))

# Low
linearMod.2 <- lm(Beginning ~ End, data = phylatest2)
print(linearMod.2)
summary(linearMod.2)

Anova(linearMod.2, type = "III")
modelSummary.2 <- summary(linearMod.2)
print(modelSummary.2)

# Place results back into a data frame
linearMod.2table <-broom::tidy(modelSummary.2)
linearMod.2table

# Save chart with p-value
saveRDS(linearMod.2table, file = here ("././results/linearMod_2table.rds"))





```









