---
title: "Supplemental Material"
subtitle: ""
author: Mia Carmichael
institute: "University of Georgia"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: article
output:
  bookdown::word_document2: 
    toc: false
bibliography: ../references.bib
csl: ../apa.csl
---

```{r, echo=FALSE, warning=FALSE}
library(dplyr)
library(pander)
library(knitr)
library(here)
library(car)
library(ggplot2)
library(broom)
library(caret)
library(vip)
library(jtools)
library(mlr)
library(ggstance)
library(huxtable)
library(readr)
library(tidyverse)
library(base)





phyladata<- readRDS(here("./data/processed_data/processeddata_phyladata.rds"))
phylatest <- readRDS(here("./data/processed_data/processeddata_phylatest.rds"))
phylacomb <- readRDS(here("././results/phylacomb.rds"))

phyladata2<- readRDS(here("./data/processed_data/processeddata_phyladata2.rds"))
phylatest2 <- readRDS(here("./data/processed_data/processeddata_phylatest2.rds"))
phylacomb2 <- readRDS(here("././results/phylacomb2.rds"))

familydata<- readRDS(here("./data/processed_data/processeddata_familydata.rds"))
familytest <- readRDS(here("./data/processed_data/processeddata_familytest.rds"))


familydata2<- readRDS(here("./data/processed_data/processeddata_familydata2.rds"))
familytest2 <- readRDS(here("./data/processed_data/processeddata_familytest2.rds"))


genusdata<- readRDS(here("./data/processed_data/processeddata_genusdata.rds"))
genustest <- readRDS(here("./data/processed_data/processeddata_genustest.rds"))


genusdata2<- readRDS(here("./data/processed_data/processeddata_genusdata2.rds"))
genustest2 <- readRDS(here("./data/processed_data/processeddata_genustest2.rds"))


```




## Bivariate Analysis
Bacterial abundance verses time period. ANOVA allows us to see that there is variance between the bacterial abundance at beginning and end of the feedlot period in High-RFI steers (Table 5, Table 6). This indicates that the microbiome did experieience changes over the course of the feedlot period.

Looking at the correlation, we can see that there is a high positive correlation between the change from the beginning of the feedlot period to the end of the feeddlot period for phyla,  and therefore the correlation between them will be closer to 1.

# Looking at ANOVA
```{r, warning=FALSE, echo=FALSE}
# High
plot(Abundance ~ Phyla, data = phylacomb)
phylahigh.aov <- aov(Abundance ~ Phyla, data = phylacomb)
summary(phylahigh.aov)
phylahigh.aovsummary <- summary(phylahigh.aov)
print(phylahigh.aovsummary)

saveRDS(phylahigh.aovsummary, file = here("./results/phylahigh_aov_table.rds"))



# Low
plot(Abundance ~ Phyla, data = phylacomb2)
phylalow.aov2 <- aov(Abundance ~ Phyla, data = phylacomb2)
summary(phylalow.aov2)
phylalow.aovsummary2 <- summary(phylalow.aov2)
print(phylalow.aovsummary2)

saveRDS(phylalow.aovsummary2, file = here("./results/phyla_low_aov_table.rds"))




```


# Looking at a correlation between data points at the beginning and end of feedlot period
```{r, warning=FALSE, echo=FALSE}

#High
library(flextable)
phylacor <- cor(phylatest$Beginning, phylatest$End)

phylacor.table <-broom::tidy(phylacor)
phylacor.table
saveRDS(phylacor.table, file = here("./results/phylahightablecor.rds"))


#Low
phylacor2 <- cor(phylatest2$Beginning, phylatest2$End)

phylacor2.table <-broom::tidy(phylacor2)
phylacor2.table
saveRDS(phylacor2.table, file = here("./results/phylahightablecor2.rds"))


```


# Looking at Linear Model
```{r, echo=FALSE, warning=FALSE}

# High
linearMod.1 <- lm(Beginning ~ End, data = phylatest)
print(linearMod.1)
summary(linearMod.1)

Anova(linearMod.1, type = "III")
modelSummary.1 <- summary(linearMod.1)
print(modelSummary.1)

# Place results back into a data frame
linearMod.1table <-broom::tidy(modelSummary.1)
linearMod.1table

# Save chart with p-value
saveRDS(linearMod.1table, file = here ("././results/linearMod_1table.rds"))

# Low
linearMod.2 <- lm(Beginning ~ End, data = phylatest2)
print(linearMod.2)
summary(linearMod.2)

Anova(linearMod.2, type = "III")
modelSummary.2 <- summary(linearMod.2)
print(modelSummary.2)

# Place results back into a data frame
linearMod.2table <-broom::tidy(modelSummary.2)
linearMod.2table

# Save chart with p-value
saveRDS(linearMod.2table, file = here ("././results/linearMod_2table.rds"))





```



## Full Analysis
# K- means Cluster Analysis

Clustering helps find subgroups of observations within a data set. By clustering observations, the desired outcome is to have observations in the same group to be similar and observations in different groups to be dissimilar. Therefor, this method seeks to find relationships between the observations without being trained by a response variable. Finally clustering allows the idenification of which observations are alike and further categorize them. 

```{r, echo=FALSE, warning=FALSE}
#### Cluster
library(tidyverse)
library(factoextra)
library(FactoMineR)
library(cluster)
```

# Phyla
```{r, phyladata,phyladata2, echo=FALSE, warning=FALSE}
#Combine low and high RFI- Phyla
phyla.all <- rbind(phyladata, phyladata2)
phyla.all
```

```{r, phyla.all, echo=FALSE, warning=FALSE}
#Rows are observations (individuals) and columns are variables
sapply(phyla.all, class)
phyla.all1=subset(phyla.all, select = -c(Phyla))
attr(phyla.all1, "row.names") <- c("H-Firmicutes", "H-Bacteroidetes", "H-Actinobacteria", "H-Proteobacteria", "H-Euryarchaeota", "H-TM7", "H-Spirochaetes", "H-Tenericutes", "H-Cyanobacteria", "H-Planctomycetes", "H-OtherPhyla", "L-Firmicutes", "L-Bacteroidetes", "L-Actinobacteria", "L-TM7", "L-Proteobacteria", "L-Euryarchaeota", "L-Spirochaetes", "L-Cyanobacteria", "L-Tenericutes", "L-Planctomycetes", "L-OtherPhyla")
view(phyla.all1)

```


```{r, echo=FALSE, warning=FALSE}
#Don’t want the clustering algorithm to depend to an arbitrary variable unit, we start by scaling/standardizing the data using the R function scale:

phyla.all1 <- scale(phyla.all1)
head(phyla.all1)
saveRDS(phyla.all1, file= here("././results/phyla.all1.rds"))
```

Clutering Distance Measures 
get_dist: for computing a distance matrix between the rows of a data matrix. 
fviz_dist: for visualizing a distance matrix

```{r, phyla.all1, echo=FALSE, warning=FALSE}
phyladistance <- get_dist(phyla.all1) 
saveRDS(phyladistance, file= here("././results/phyladistance.rds"))
D1<-fviz_dist(phyladistance, gradient = list(low= "#00AFBB", mid= "white", high= "#FC4E07")) + ggtitle("Phyla Distance Matrix")
```


Assessing Clusterability
```{r, echo=FALSE, warning=FALSE}
library(factoextra)
library(ggplot2)
#The function get_clust_tendency() [factoextra package] can be used. It computes the Hopkins statistic
phyla.allclust <- get_clust_tendency(phyla.all1, 21, graph = TRUE)
phyla.allclust$hopkins_stat
print(phyla.allclust$plot)
ggsave(filename = here("./results/phylaclusterdistance.png"), plot = phyla.allclust$plot)

```
The value of the Hopkins statistic is significantly < 0.5 (0.94368), indicating that the data is highly clusterable. Additionally, It can be seen that the ordered dissimilarity image contains patterns (i.e., clusters).

Estimate the number of clusters in the data, method 1
```{r, echo=FALSE, warning=FALSE}

#k-means clustering requires to specify the number of clusters to generate, use the function clusGap() to compute gap statistics for estimating the optimal number of clusters . The function fviz_gap_stat() is used to visualize the gap statistic plot.


#Compute gap statistic
set.seed(22)
phyla.gap <- clusGap(phyla.all1, FUN = kmeans, nstart = 25, K.max = 10, B = 100)

#Plot the result
fviz_gap_stat(phyla.gap)

#The gap statistic suggests a 7 cluster solution

#Compute k-means clustering, k=7
set.seed(22)
km.phyla <- kmeans(phyla.all1, 7, nstart = 25)
head(km.phyla$cluster, 22)

#Visualize clusters using factoextra
fviz_cluster(km.phyla, phyla.all1)


#Cluster validation statistics: Inspect cluster silhouette plot
#Silhouette measures (Si) how similar an object i is to the the other objects in its own cluster versus those in the neighbor cluster. Si
#Si values range from 1 to - 1:
sil.phyla <- silhouette(km.phyla$cluster, dist(phyla.all1))
rownames(sil.phyla) <- rownames(phyla.all1)
head(sil.phyla)
fviz_silhouette(sil.phyla)
  
#It can be seen that there are some samples which have negative silhouette values. Some natural questions are : "Which samples are these? To what cluster are they closer?"
neg.sil.phyla <- which(sil.phyla[,"sil_width"] < 0)
sil.phyla[neg.sil.phyla, , drop = FALSE]

```

K-means clustering using eclust(), method 2 

```{r, echo=FALSE, warning=FALSE}

#K-means 
K1<-eclust(phyla.all1, "kmeans", nstart = 25)
saveRDS(K1, file= here("././results/K1.rds"))
#Gap statistic plot
G1 <- fviz_gap_stat(K1$gap_stat) + ggtitle("Phyla Gap Statistic Plot")

# Silhouette plot
S1 <-fviz_silhouette(K1) + ggtitle("Phyla Silhouette Plot")

#Enhanced hierarchical clustering

hc.phyla <- eclust(phyla.all1, "hclust")
saveRDS(hc.phyla, file= here("././results/hc.phyla.rds"))

H1 <- fviz_dend(hc.phyla, rect = TRUE) + ggtitle("Phyla Hierarchial Clustering")

#Silhouette plot and the scatter plot for hierarchical clustering.

SH1 <- fviz_silhouette(hc.phyla) + ggtitle("Phyla Hierarchial Silhouette Plot")
SPH1 <- fviz_cluster(hc.phyla) + ggtitle("Phyla Hierchial Scatter Plot")


saveRDS(K1, file = here("././results/K1.rds"))
saveRDS(G1, file =here("./results/G1.rds"))
saveRDS(H1, file = here("./results/H1.rds"))
saveRDS(SH1, file = here("./results/SH1.rds"))
saveRDS(SPH1, file = here("./results/SPH1.rds"))
saveRDS(S1, file = here("./results/S1.rds"))


```


Choose a method and do for all (phyla, family, and genus)
# Family
```{r, echo=FALSE, warning=FALSE}
#Combine low and high RFI- Family
family.all<- rbind(familydata, familydata2)

#Rows are observations (individuals) and columns are variables
sapply(family.all, class)
family.all1 = subset(family.all, select = -c(`Family`))
attr(family.all1, "row.names") <- c("H-Ruminococcaceae", "H-Prevotellaceae", "H-Order Clostridiales", "H-Order Bacteroidales", "H-Lachnospiraceae", "H-RF16", "H-Methanobacteriaceae", "H-F16", "H-Paraprevotellaceae", "H-S24-7", "H-Coriobacteriaceae", "H-Erysipelotrichaceae", "H-Bifidobacteriaceae", "H-Spirochaetaceae", "H-Streptococcaceae", "H-Mogibacteriaceae", "H-Succinivibrionaceae", "H-Other Families" , "L-Ruminococcaceae", "L-Prevotellaceae", "L-Order Bacteroidales", "L-Order Clostridiales", "L-Lachnospiraceae", "L-Bifidobacteriaceae", "L-F16", "L-RF16", "L-Paraprevotellaceae", "L-Methanobacteriaceae", "L-Spirochaetaceae", "L-Erysipelotrichaceae", "L-Coriobacteriaceae", "L-Succinivibrionaceae", "L-Order RF32", "L-S24-7", "L-Other Families")
view(family.all1)
```


```{r, echo=FALSE, warning=FALSE}
#Don’t want the clustering algorithm to depend to an arbitrary variable unit, we start by scaling/standardizing the data using the R function scale:

family.all <- scale(family.all1)
head(family.all1)
saveRDS(family.all1, file= here("././results/family.all1.rds"))
```


```{r, echo=FALSE, warning=FALSE}
#Assessing Clusterability
#The function get_clust_tendency() [factoextra package] can be used. It computes the Hopkins statistic
family.all.hop <- get_clust_tendency(family.all1, 34, graph = TRUE)

family.all.hop$hopkins_stat

#The value of the Hopkins statistic is significantly < 0.5 (0.9490412), indicating that the data is highly clusterable. Additionally, It can be seen that the ordered dissimilarity image contains patterns (i.e., clusters).
```


```{r, echo=FALSE, warning=FALSE}
#K-means clustering using eclust(). Doesnt allow duplicate row names, will have to change. 

K2 <-eclust(family.all1, "kmeans", nstart = 25)
saveRDS(K2, file= here("././results/K2.rds"))

#Gap statistic plot
G2 <-fviz_gap_stat(K2$gap_stat)  + ggtitle("Family Gap Statistic Plot")


# Silhouette plot
S2 <- fviz_silhouette(K2) + ggtitle("Family Silhouette Plot")


#Enhanced hierarchical clustering

hc.family <- eclust(family.all1, "hclust")
saveRDS(hc.family, file= here("././results/hc.family.rds"))

H2 <-fviz_dend(hc.family, rect = TRUE) + ggtitle("Family Hierarchial Clustering")

#Silhouette plot and the scatter plot for hierarchical clustering.

SH2 <- fviz_silhouette(hc.family) + ggtitle("Family Hierarchial Silhouette Plot")
SPH2 <- fviz_cluster(hc.family) + ggtitle("Family Hierarchial Scatter Plot")




saveRDS(K2, file = here("././results/K2.rds"))
saveRDS(G2, file =here("./results/G2.rds"))
saveRDS(H2, file = here("./results/H2.rds"))
saveRDS(SH2, file = here("./results/SH2.rds"))
saveRDS(SPH2, file = here("./results/SPH2.rds"))
saveRDS(S2, file = here("./results/S2.rds"))

```



# Genus

```{r, echo=FALSE, warning=FALSE}
#Combine low and high RFI- Genus
genus.all <- rbind(genusdata, genusdata2)

#Rows are observations (individuals) and columns are variables
sapply(genus.all, class)
genus.all1 = subset(genus.all, select = -c(`Genus`))
attr(genus.all1, "row.names") <- c("H-Methanobrevibacter","H-Methanosphaera","L-Methanobrevibacter", "L-Methanosphaera")
view(genus.all1)

#Don’t want the clustering algorithm to depend to an arbitrary variable unit, we start by scaling/standardizing the data using the R function scale:

genus.all1 <- scale(genus.all1)
head(genus.all1)
saveRDS(genus.all1, file= here("././results/genus.all1.rds"))


#Assessing Clusterability
#The function get_clust_tendency() [factoextra package] can be used. It computes the Hopkins statistic
genus.all.hop <- get_clust_tendency(genus.all1, 3, graph = TRUE)

genus.all.hop$hopkins_stat

#The value of the Hopkins statistic is not significant > 0.5 (0.4842), indicating that the data is not highly clusterable. Additionally, It can be seen that the ordered dissimilarity image does not contains patterns (i.e., clusters).


#K-means clustering using eclust(). 

K3 <-eclust(genus.all1, "kmeans", nstart = 25, k.max = 2 )
saveRDS(K3, file= here("././results/K3.rds"))

#Gap statistic plot
G3 <-fviz_gap_stat(K3$gap_stat)  + ggtitle("Genera Gap Statistic Plot")


# Silhouette plot: was unable to produce silhouette plot for genera. 


#Enhanced hierarchical clustering

hc.genus <- eclust(genus.all1, "hclust", k = 2)
saveRDS(hc.genus, file= here("././results/hc.genus.rds"))

H3 <-fviz_dend(hc.genus, rect = TRUE) + ggtitle("Genera Hierarchial Clustering")


#Silhouette plot and the scatter plot for hierarchical clustering.

SH3 <- fviz_silhouette(hc.genus) + ggtitle("Genera Hierarchial Silhouette Plot")
SPH3 <- fviz_cluster(hc.genus) + ggtitle("Genera Hierarchial Scatter Plot")




saveRDS(K3, file = here("././results/K3.rds"))
saveRDS(G3, file =here("./results/G3.rds"))
saveRDS(H3, file = here("./results/H3.rds"))
saveRDS(SH3, file = here("./results/SH3.rds"))
saveRDS(SPH3, file = here("./results/SPH3.rds"))







```

# Combine and save figures 
```{r, echo=FALSE, warning=FALSE}

library(gridExtra)
#Compile all Kmeans figures for comparison
PK1 <- fviz_cluster(K1, geom = "point", data = df) + ggtitle("Phyla K-Means: k = 3")
PK2 <- fviz_cluster(K2, geom = "point",  data = df) + ggtitle("Family K-Means: k = 7")
PK3 <- fviz_cluster(K3, geom = "point",  data = df) + ggtitle("Genra K-Means: k = 1")

library(gridExtra)
all.km <-grid.arrange(PK1, PK2, PK3, nrow = 3) 

#Compile all gap figures for comparison
all.gap <- grid.arrange(G1,G2,G3, nrow = 3)

#Compile all Silhouette figures for comparison
all.sil <- grid.arrange(S1,S2, nrow = 2)
                        
#Compile all Hierarchial Clustering figures for comparison
all.hier <- grid.arrange(H1, H2, H3, nrow = 3)

#Compile all Hierarchial Silhouette figures for comparison
all.hsil <- grid.arrange(SH1, SH2, SH3, nrow = 3)

#Compile all Hierarchial Scatter Plot figures for comparison
all.hsp <- grid.arrange(SPH1, SPH2, SPH3, nrow = 3)

#Save 

saveRDS(all.km, file = here("././results/all.km.rds"))
saveRDS(all.gap, file = here("././results/all.gap.rds"))
saveRDS(all.sil, file = here("././results/all.sil.rds"))
saveRDS(all.hier, file = here("././results/all.hier.rds"))
saveRDS(all.hsil, file = here("././results/hsil.rds"))
saveRDS(all.hsp, file = here("././results/all.hsp.rds"))


ggsave(filename = here("././results/all.km.png"),plot=all.km)
ggsave(filename = here("././results/all.gap.png"),plot=all.gap)
ggsave(filename = here("././results/all.sil.png"),plot=all.sil)
ggsave(filename = here("././results/all.hier.png"),plot=all.hier)
ggsave(filename = here("././results/all.hsil.png"),plot=all.hsil)
ggsave(filename = here("././results/all.hsp.png"),plot=all.hsp)

```



```{r}

#Tidy data to run analysis

#Combine data
F1 <- rbind(familydata, familydata2)
F1

#Create new column with change in bacterial abundance over time
F1 <- F1 %>% mutate(Abundance.Difference = (End - Beginning))
F1 <- F1 %>% select(Family, Abundance.Difference)
F1
str(F1)

#Duplicate/create new family column in order to create Bacteria Family as a factor
F1 <- F1 %>% mutate(Bacterial.Family = Family)
str(F1)

F1$Family=as.factor(F1$Family)
F1$Family=as.numeric(F1$Family)
lapply(F1, levels)

F1$Bacterial.Family=as.factor(F1$Bacterial.Family)
lapply(F1, levels)

F1$Family=as.numeric(F1$Family)
lapply(F1, levels)

F1$Family=as.factor(F1$Family)
lapply(F1, levels)
saveRDS(F1, file = here("././results/F1.rds"))

#Create new data frame to add in RFI status
RFI_Status <- c(2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1)
CatRFI <- cut(RFI_Status, breaks = c(0,1,2), labels =c("LowRFI", "HighRFI"))

RFI_Group <- data_frame(CatRFI)
#Merge two dataframes 
F2 <-cbind(F1, CatRFI)
F2

#Change name of RFI colunm
names(F2)[1:4]<-c("Family", "Abundance_Difference", "Bacterial.Family", "RFI_Status")
F2
saveRDS(F2, file = here("././results/F2.rds"))

```



Linear Regression Model: Abundance ~RFI Status
```{r}
#Simple
#By default, the category that R chooses to be the reference/ baseline, is the first category that appears (ie. LowRFI). In LR model, the intercept refers to the estimated mean Y value for the reference group and the model coefficients, refer to expected changes in the mean Y-value relative to the reference group. 

#reference LowRFI
mod1= lm(Abundance_Difference ~ RFI_Status, data = F2)
tidy(mod1)
mod1sum <- summary(mod1)
saveRDS(mod1sum, file = here("././results/mod1sum.rds"))

#Multi
mod2=lm(Abundance_Difference ~ RFI_Status + Bacterial.Family, data = F2 )
mod2
tidy(mod2)
#To check for significant varibles 
mod2sum <- summary(mod2)
saveRDS(mod2sum, file = here("././results/mod2sum.rds"))

#Approximately 98% of variation in family abundance over time can be explained by RFI status and classification of bacterial families. 
#There are a lot of signif values and the p-value overall is small. This linear model appears to be a good fit for this data set.  


```

Plotting Linear Regression Model
```{r}
#Single model

plot_summs(mod1)

#Differing scales of the varibles 
plotmod2 <- plot_summs(mod2, scale = TRUE)
saveRDS(plotmod2, file = here("././results/plotmod2.rds"))

#The default width of the confidencfe interval is .95
plotmod2sum <- plot_summs(mod2, scale = TRUE, inner_ci_level = .9)
saveRDS(plotmod2sum, file = here("././results/plotmod2sum.rds"))

#Plot coefficient uncertainty as normal distributions
plotmod2sum2 <- plot_summs(mod2, scale = TRUE, plot.distributions = TRUE, inner_ci_level = .9)

saveRDS(plotmod2sum2, file = here("././results/plotmod2sum2.rds"))
```


Table Outputts for Word/RMD
```{r}
expmod2 <- export_summs(mod2, scale = TRUE)


#add confidence intervals 
expmod2.2 <- export_summs(mod2, scale = TRUE, error_format = "[{conf.low}, {conf.high}]")


#add p values
expmod2.3 <- export_summs(mod2, scale = TRUE, error_format = "[{statistic}, p = {p.value}]")
saveRDS(expmod2.3, file = here("././results/expmod2.3.rds"))
```


Looking at other models: Logistic Regression Model
```{r}

#Change RFI_Status baack to numbers 
levels(F2$RFI_Status) <- c("1", "2")
lapply(F2, levels)
F2


#RFI_Status ~ Abundance_Difference
mod3 <- glm(RFI_Status ~ Abundance_Difference, family = "binomial", data = F2)
tidy(mod3)
mod3sum <- summary(mod3)
saveRDS(mod3sum, file = here("././results/mod3sum.rds"))



#RFI_Status ~ Family
mod4 <- glm(RFI_Status ~ Family, family = "binomial", data = F2)
tidy(mod4)
summary(mod4)
mod4sum <- summary(mod4)
saveRDS(mod4sum, file = here("././results/mod4sum.rds"))


#RFI_Status ~ Family + Abundance_Difference +Bacterial. Family
mod5 <- glm(RFI_Status ~ Family + Abundance_Difference + Bacterial.Family, family = "binomial", data = F2)
tidy(mod5)
summary(mod4)
```









